# Data Vizualization

One of the strengths of R is its capability for data visualization. While base R itself offers robust visualization and graphing capabilities, the open-source nature of R encourages the development and use of various packages to extend its visualization potential. Among these, the `ggplot2::` package stands out for its powerful and versatile graphing functions, making it a popular choice among data scientists and statisticians.

Although R is a programming language with a syntax that can be complex for beginners, the `esquisse::` package provides a user-friendly interface to facilitate graph plotting. By simplifying the process of creating visualizations, `esquisse::` allows users to generate plots more intuitively, making data visualization accessible to those less familiar with R's syntax.

In this chapter, we will demonstrate examples of several types of graphs, showcasing the breadth of R's visualization capabilities. Through practical examples, readers will learn how to leverage packages like `ggplot2::` and tools like `esquisse::` to transform data into compelling visual narratives. Whether you're visualizing trends, distributions, relationships, or comparisons, R provides the tools necessary to produce informative and aesthetically pleasing graphics.

Some of the commong graph here are

-   Single Variable
    -   histogram
    -   boxplot
    -   bar chart
    -   scatter plot
    -   line chart
    

## Practical - Setup Project

1.  Setup your project
    -   Open your RStudio
    -   Create New Project
2.  Create Quarto document
    -   update the YAML metadata to make the document self-contained

```{r}
#| echo: true
#| eval: false

---
title: "Sesi 2 - Data Vizualization"
format:
  html:
    embed-resources: true
---

```

3.  Download the dataset
    -   we will use `asthmads_clean.sav` data set


```{r}
#| echo: fenced
#| eval: false

download.file(
  url = "https://github.com/MohdAzmiSuliman/IKU_RBook/raw/master/dataset/asthmads_clean.sav", 
  destfile = "asthmads_clean.sav", mode = "wb")
```


```{r}
#| echo: true

library(tidyverse)
library(haven)

asthmads <- read_sav("asthmads_clean.sav") %>% 
  as_factor()
asthmads
```


## `esquisse::` RStudio Addin

## Histogram

Histogram suitable to show distribution of single numerical variable, such as the body weight distribution

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(Weight_Pre)) + 
  geom_histogram()
```


we can customise it to make it prettier

```{r}
asthmads %>% 
  ggplot(aes(Weight_Pre)) + 
  geom_histogram(binwidth = 5, alpha = 0.5, 
                 fill = "white", colour = "black") +
  theme_bw()
```


we can also stratified the graph, for example, by gender

```{r}
asthmads %>% 
  ggplot(aes(Weight_Pre)) + 
  geom_histogram(aes(fill = Gender), 
                 position = position_dodge(), 
                 binwidth = 5, alpha = 0.5, 
                 colour = "black") +
  theme_bw()
```


## Boxplot

Boxplot also a common graph use to visualize single numerical variable


```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(Weight_Pre)) + 
  geom_boxplot()
```


we can customise it to make it prettier

```{r}
asthmads %>% 
  ggplot(aes(Weight_Pre)) + 
  geom_boxplot() + 
  theme_bw()
```


we can also stratified the graph, for example, by gender

```{r}
asthmads %>% 
  ggplot(aes(Weight_Pre)) + 
  geom_boxplot(aes(fill = Gender)) + 
  theme_bw()
```


## Bar Chart

Similar to histogram, bar chart suitable to show distribution of single categorical variable, such as BMI categories. the bar in histogram is continuous, but in bar chart, has gap.

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(BMI_PreCat)) + 
  geom_bar()
```

we can customise it to make it prettier

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(BMI_PreCat)) + 
  geom_bar(aes(fill = BMI_PreCat)) + 
  theme_bw()
```

there are various way if we want to stratified this.

-   we can stratified by facet_wrap function

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(BMI_PreCat)) + 
  geom_bar(aes(fill = BMI_PreCat)) + 
  facet_wrap(~ Gender) + 
  theme_bw()
```


-   we can stratified by position dodge

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(BMI_PreCat)) + 
  geom_bar(aes(fill = Gender), position = position_dodge()) + 
  theme_bw()
```



## Scatter Plot

Scatter plot suitable to visualize two numerical variables together, for example, we want to see the relationship between body weight and height.

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(x = Ht_m, y = Weight_Pre)) + 
  geom_point()
```

we can customise it so make it pretty

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(x = Ht_m, y = Weight_Pre)) + 
  geom_point() +
  theme_bw()
```

we can also stratified by gender


```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(x = Ht_m, y = Weight_Pre)) + 
  geom_point(aes(colour = Gender)) +
  theme_bw()
```


and even add best fit line (pearson correlation)

```{r}
#| echo: fenced

asthmads %>% 
  ggplot(aes(x = Ht_m, y = Weight_Pre)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  theme_bw()
```


## Line Chart

```{r}
tibble(year = seq(2008, 2020, 4), 
       male = sample(10:25, 4, replace = T), 
       female = sample(10:25, 4, replace = T), 
       total = male+female) %>% 
  ggplot(aes(x = year, y = total)) +
  geom_point() + 
  geom_line(group = 1)
```



## Spaghetti Plot

```{r}
asthmads %>% 
  select(id, BMI_Pre, BMI_Post) %>% 
  pivot_longer(cols = c(BMI_Pre, BMI_Post)) %>% 
  ggplot(aes(x = name, y = value)) + 
  geom_point() +
  geom_line(aes(group = id))
```

